---
title: "snag-wlf_rf-mod"
author: "Jess M. Stitt"
date: "9/17/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **1. Set up the workspace**
Pre-processing and organization of materials needed to run the code, including R packages, folder & file pathways, and relevant datasets
## *1.1. Load packages for workflow organization & tidy data*
```{r library}
##--------------------------------------------------------------------------##
## Workflow organization ----
##--------------------------------------------------------------------------##
library(here)       #for file path mgmt
library(beepr)      #for [unnecessary] sound notifications
##--------------------------------------------------------------------------##
## Tidy data management ----
##--------------------------------------------------------------------------##
library(tidyverse)  #for tidy data
library(dplyr)      #for
library(tidyr)      #for
library(stringr)    #for
##--------------------------------------------------------------------------##
## Spatial manipulation ----
##--------------------------------------------------------------------------##
library(raster)
library(rgdal) #for vector work
library(raster) #for metadata/attributes: vectors or rasters
library(rgeos)
library(sf)
library(sp)
##--------------------------------------------------------------------------##
## Data visualization & plotting ----
##--------------------------------------------------------------------------##
library(ggplot2)    #for plots & graphics
library(gridExtra)
library(ggpubr)     #for ggplot-based publication-ready plots & other visuals
library(ggsci)
library(RColorBrewer)   #for color palettes
##--------------------------------------------------------------------------##
## Modeling ----
##--------------------------------------------------------------------------##
library(randomForest) #Random Forest modeling
# devtools::install_github("MI2DataLab/randomForestExplainer")
library(randomForestExplainer) #explainer for RF models
library(rfUtilities)
library(rpart)
library(rpart.plot)
library(caret)
library(corrplot)
library(rstatix)
library(rsample)
##--------------------------------------------------------------------------##
beep(1) #----
```
## *1.2. Assign project file pathways*
```{r pathways}
##--------------------------------------------------------------------------##
## INPUTS: existing datasets ----
##--------------------------------------------------------------------------##
csv <- here("01_datasets", "01_csv")    #field data files (as comma-separated)
laspc <- here("01_datasets", "02_las")  #lidar point cloud files
gis <- here("01_datasets", "03_gis")    #geospatial data files
##--------------------------------------------------------------------------##
## PROCESSING: R scripts and steps for project ----
##--------------------------------------------------------------------------##
### Code used to perform analyses
scripts <- here("02_scripts")           #R scripts
### Lidar progression
rawlas <- here(laspc,"00_las_raw_r50m")   
znf <- here(laspc,"01_las_znfil_r50m")       #znorm filtered (>0, C==1L) R50m
clipPlot <- here(laspc,"02_las_clip-plot_r25m") #lasclipCircle from znf to R25m
clipTree <- here(laspc,"03_las_clip-tree") #lasclipCircle from znf to 5x5m
clipSnag <- here(clipTree,"circular_d05m/snag-gnss") #field-derived snag PCs
### Raster progression
dtm50m <- here(gis,"00_tif_dtm_r50m")
pfc <- here(gis,"01_tif_pf-chm_r50m")   #pit-free chm R50m around plots
chm50m <- here(gis,"02_asc_pf-chm_r50m")      #chm raster R50m around plots
chmPlot <- here(gis,"03_asc_chm-plot_r25m")       #chm raster from chm to R25m
shp25m <- here(gis, "04_shp_plots-clip_r25m")
##--------------------------------------------------------------------------##
## OUTPUTS: products generated during processing ----
##--------------------------------------------------------------------------##
tabs <- here("03_results", "01_tables") #results tables (as .csvs)
lasmx <- here("03_results", "02_las-metrics") #lidar-derived metrics
shp <- here("03_results", "03_gis-mod")  #shapefiles (modified)
figs <- here("03_results", "04_graphics") #graphic outputs
##--------------------------------------------------------------------------##
beep(2)  #----
##--------------------------------------------------------------------------##
```

# **2. ALS processing**
Using the height-normalized airborne laser scanning (ALS) point clouds with a 50m-radius, centered on the survey plot centers, all snags per 25m-radius survey plot were clipped

[NOTE: for this project, the lidar data has not been included but the code shows how the ALS point clouds were processed and the specific lidar metrics were calculated]
## *2.1. Clip LAS around individual snags*
```{r lasproc}
##--------------------------------------------------------------------------##
## Lidar-specific packages ----
##--------------------------------------------------------------------------##
## **If not yet installed, first install with devtools
# library(devtools) #for pkg maintenance [developer tools]
# devtools::install_github("carlos-alberto-silva/rLiDAR")
library(rLiDAR)     #for lidar data manipulation & ITD**
# devtools::install_github("Jean-Romain/lidR")
library(lidR)     #for lidar data manipulation**
##--------------------------------------------------------------------------##
##--------------------------------------------------------------------------##
## Clip height-normalized las point clouds around each snag (size TBD) ----
##--------------------------------------------------------------------------##
list_snagClips_d05m <- vector("list", nsnags) 
for(i in 1:length(snagNames)){
    plotZNF_tmp <- paste0(znf, "/",  #file path to las for plot ID in snag i
                          str_sub(snagNames[[i]], end = -6), "_znf.las")
    plotZNF <- lidR::readLAS(plotZNF_tmp)   #read in znorm r50m point cloud
    tmpSNAG <- lidR::clip_circle(plotZNF,
                                 snag_chars$xc[[i]],
                                 snag_chars$yc[[i]],
                                 radius = 2.5)
    list_snagClips_d05m[[i]] <- tmpSNAG
    writeLAS(tmpSNAG, paste0(clipSnag, "/",
                             str_sub(snagNames[[i]]), ".las")) #write to file
}; beep(2)
names(list_snagClips_d05m) <- paste0(snagNames) #include snag IDs for ref
##--------------------------------------------------------------------------##
## Catalog and inspect created files ----
##--------------------------------------------------------------------------##
ctgSNAG <- readLAScatalog(clipSnag) #catalog all files in folder
lidR:::catalog_laxindex(ctgSNAG)    #index all files & create lax files
las_check(ctgSNAG)                  #check files
plot(ctgSNAG, mapview=T)            #view clips on a map
##--------------------------------------------------------------------------##
beep(1) #----
##--------------------------------------------------------------------------##
```
## *2.2. Calculate lidar metrics from individual snag point clouds*
```{r lasmx}
##--------------------------------------------------------------------------##
## Calculate "standard" lidar metrics (via lidR) for all returns ----
##--------------------------------------------------------------------------##
## Create list of all clips of lidar point clouds around snags
snagLAS_list <- list.files(paste0(clipSnag, "/"), pattern="*.las")
## Iterate through list: read, process, calculate metrics
snagMx_lidr <- lapply(snagLAS_list, function(file) {
    las <- lidR::readLAS(paste0(clipSnag,"/",file))
    las <- filter_poi(las, Z>=1.37) #set min.ht to breast height (dbh)
    lidR::cloud_metrics(las, .stdmetrics)
})
snagMx_d <- data.table::rbindlist(snagMx_lidr)
snagMx_t <- as_tibble(cbind(snagid=c(snagNames), snagMx_d)) %>%
    select(-c(38:49)) %>% #remove intensity metrics... not calibrated
    mutate(ptd = n / 25, #standardize to point density (=returns per meter)
           zcv = zmean / zsd)
glimpse(snagMx_t)
## Save to file
# write_csv(snagMx_t, file.path(lasmx,
#                               "snag-trem_REFsnags-d05m_cloudMx-allRN.csv"))
##--------------------------------------------------------------------------##
## Set up additional structural metrics, altered from existing methods ----
##--------------------------------------------------------------------------##
## Calculate gap fraction profile (derived from lidR::gap_fraction_profile):
gap_frac_prof_jstitt <- function (z, dz = 1, z0 = 1) 
{
    bk <- seq(z0,
              ceiling(max(z) + z0), 
              dz)
    if (length(bk) <= 1) 
        return(data.frame(z = numeric(0), gf = numeric(0)))
    histogram <- graphics::hist(z, breaks = bk, plot = F)
    h <- histogram$mids
    p <- histogram$counts/sum(histogram$counts)
    p <- c(p, 0)
    cs <- cumsum(p)
    i <- data.table::shift(cs)/cs
    i[is.na(i)] = 0
    i[is.nan(i)] = NA
    z = h
    i = i[-length(i)]
    return(na.omit(data.frame(z = z[z > z0], gf = i[z > z0])))
}
## Calculate structural methods (derived from LaRue et al. 2020**): 
##   ** Based on a tutorial found at: https://www.neonscience.org/resources/
##      learning-hub/tutorials/structural-diversity-discrete-return
strMx_jstitt <- function(lasroi) {
    Zs <- tmpLAS@data$Z
    Zs <- Zs[!is.na(Zs)]
    if(length(Zs)>28) {
    chm <- grid_canopy(lasroi, res = 1, dsmtin())
    rumple <- rumple_index(chm)
    top.rugosity <- sd(chm@data@values, na.rm = TRUE)
    cells <- length(chm@data@values)
    chm.0 <- chm
    chm.0[is.na(chm.0)] <- 0
    zeros <- which(chm.0@data@values == 0)
    deepgaps <- length(zeros)
    deepgap.fraction <- deepgaps/cells
    cover.fraction <- 1 - deepgap.fraction
    entropy_mid <- entropy(Zs, by = 1, zmax = 20) 
    gap_frac <- gap_frac_prof_jstitt(Zs, dz = 1, z0 = 1.37)
    GFP <- mean(gap_frac$gf, na.rm = TRUE)
    GFP_mid <- mean(gap_frac$gf[gap_frac$z <= 20])
    LADen <- LAD(Zs, dz = 2, k=0.5, z0=2) 
    LADm <- mean(LADen$lad, na.rm=TRUE)
    LADsd <- sd(LADen$lad, na.rm=TRUE)
    LADcv <- LADm/LADsd
    VAI <- sum(LADen$lad, na.rm=TRUE) 
    VCI <- VCI(Zs, by = 1, zmax=80) 
    out.plot <- data.frame(
        matrix(c(rumple, top.rugosity, deepgap.fraction, cover.fraction, 
                 entropy_mid, GFP, GFP_mid,
                 LADm, LADcv, VAI, VCI),
            ncol = 11)) 
    colnames(out.plot) <- 
        c("rumple", "top.rugosity", "deepgap.fraction", "cover.fraction",
          "ENTmid", "GFPall", "GFPmid",
          "LADm", "LADcv", "VAI", "VCI") 
    print(out.plot)
    } else {
        vals <- rep(NA, times = 11)
        out.plot <- data.frame(
            matrix(vals, ncol=11, nrow=1, byrow=FALSE))
    colnames(out.plot) <- 
        c("rumple", "top.rugosity", "deepgap.fraction", "cover.fraction",
          "ENTmid", "GFPall", "GFPmid",
          "LADm", "LADcv", "VAI", "VCI") 
    print(out.plot)
    }
}
##--------------------------------------------------------------------------##
## Calculate structural metrics for all returns ----
##--------------------------------------------------------------------------##
strMetrics_allRN = list()
for(i in 1:length(snagNames)){
    snagLAS_tmp <- paste0(clipSnag, "/", 
                          str_sub(snagNames[[i]]), ".las")
    tmpLAS <- lidR::readLAS(snagLAS_tmp, 
                            filter = "-drop_z_below 1.37")
    tmpMX <- strMx_jstitt(tmpLAS) 
    strMetrics_allRN[[i]] <- tmpMX
}
snagStrMx_allRN <- data.table::rbindlist(strMetrics_allRN)
snagMx_str <- as_tibble(cbind(snagid=c(snagNames), snagStrMx_allRN)) %>%
    tidyr::drop_na(rumple)
glimpse(snagMx_str)
## Save to file
# write_csv(snagMx_str, file.path(lasmx, 
#                                 "snag-trem_REFsnags-d05m_structMx-allRN.csv"))
##--------------------------------------------------------------------------##
## Combine all metrics and snag characteristics for evaluation ----
##--------------------------------------------------------------------------##
## Combine standard & structural metrics
snagStr_returns <- full_join(snagMx_t, snagMx_str, by="snagid")
## Combine all metrics with snag characteristics (from field data)
snag_lasmx <- inner_join(snag_chars, snagStr_returns, by="snagid") %>% 
    drop_na(c(dbh,top)) 
snagMx <- snag_lasmx %>%
    select(-c(3:5,9,10,63)) %>%
    tidyr::drop_na(rumple); snagMx
## Save to file
# write_csv(snagMx, file.path(lasmx,
#                             "snag-trem_REFsnags-d05m_lasMx-allRN_full.csv"))
##--------------------------------------------------------------------------##
beep(3) #----
##--------------------------------------------------------------------------##
```
## *2.3. Add in topographic information clipped from DEM for each snag*
```{r demtopo}
##--------------------------------------------------------------------------##
## Read in raw PCs to generate DTMs for topo (R50m) ----
##--------------------------------------------------------------------------##
ctgROI <- lidR::readLAScatalog(rawlas)    #catalog raw las, R50m plots
opt_independent_files(ctgROI) <- TRUE
lidR:::catalog_laxindex(ctgROI)     #index all files & create lax files
las_check(ctgROI)
##--------------------------------------------------------------------------##
## DEM for REF plots (R50m) ----
##--------------------------------------------------------------------------##
opt_output_files(ctgROI) <- (paste0(dtm50m,"/{*}_dtm"))
ctgDTM <- lidR::grid_terrain(ctgROI, 20, algorithm = tin())
opt_independent_files(ctgDTM) <- TRUE
##--------------------------------------------------------------------------##
## Use raster::terrain to calculate slope and aspect ----
##--------------------------------------------------------------------------##
dtm_slope <- terrain(ctgDTM, opt = "slope")
dtm_aspect <- terrain(ctgDTM, opt = "aspect", unit="degrees")
##--------------------------------------------------------------------------##
## Extract topo features at survey plot scale (circular 25m radius) ----
##--------------------------------------------------------------------------##
## Average slope
avgslope <- raster::extract(dtm_slope,
  sp, 
  buffer=25,
  fun=mean,
  na.rm=TRUE, 
  df=TRUE)
## Convert degrees of slope into percentage of slope: 
#   Tan(Slope-in-degrees * Pi/180)
## Convert slope in percentage to slope in degrees:
#   Atan(slope-in-percent)*180/Pi

## Average aspect
avgasp <- raster::extract(dtm_aspect,
  sp,
  buffer=25,
  fun=mean, 
  na.rm=TRUE, 
  df=TRUE)

## Transformed aspect (TrAsp)
trasp <- function(x) { (1 - cos( (pi/180) * (x - 30)) ) / 2 }
options("scipen"=10, "digits"=2)
dtm_trasp <- as_tibble(trasp(avgasp$aspect)) %>% 
  rename(trasp = value)

## Also extract elevation averaged at the plot scale (circular 25m radius)
avgelev <- raster::extract(ctgDTM,
  sp,
  buffer=25,
  fun=mean, 
  na.rm=TRUE,
  df=TRUE)  
##--------------------------------------------------------------------------##
## Combine all topo features ----
##--------------------------------------------------------------------------##
tidytopo <- dtm_trasp %>% 
  add_column(plotid = plots$plotid, .before = "trasp") %>% 
  add_column(as_tibble(avgslope)) %>% select(-"ID") %>% 
  add_column(as_tibble(avgelev), .before = "trasp") %>% 
  select(-"ID") %>% 
  rename(elev = BR210A_dtm)
## Save file
# write_csv(tidytopo, file.path(tabs,
#                               "snag-trem_REFplots_topo-r25m.csv"))
##--------------------------------------------------------------------------##
beep(10) #----
##--------------------------------------------------------------------------##
```

# **3. Random Forest Modeling**
## *3.1. Read in datasets (if not processed in situ per section 2 above)*
```{r readr}
##--------------------------------------------------------------------------##
## Load relevant csv files ----
##--------------------------------------------------------------------------##
## Load csv with all lidar metrics and snag characteristics
snagDF <- read_csv(file.path(lasmx,
                             "snag-trem_REFsnags-d05m_lasMx-allRN_full.csv"))
## Modify for model processing
snagMOD <- as_tibble(snagDF) %>% 
  mutate_at(vars(site, spp, top:insects), factor) %>%
  reorder_levels(bark, c("NONE","SOME","FULL")) %>%
  reorder_levels(branches, c("N","C","M","F","V")) %>% #None,Coarse,Mid,Fine,Veg
  mutate(overcanopy = zmax - ht) %>%
  mutate(cat=cut(dbh, breaks=c(0, 0.4, 2), 
                 labels=c("small","large"))) %>%
  mutate(group = case_when(
    cat == "small" & top == "INTACT" ~ "SI",
    cat == "large" & top == "INTACT" ~ "LI",
    cat == "small" & top == "BT" ~ "SB",
    cat == "large" & top == "BT" ~ "LB"),
    .after=site) %>%
  mutate(crown = case_when(
    branches == "N" | branches == "C" ~ "NO",
    branches == "M" | branches == "F" | branches == "V" ~ "YES"),
    .after=group) %>%
  mutate_at(vars(group, crown), factor) %>% 
  rstatix::reorder_levels(group, c("SI","LI","SB","LB"))
## Add column on plot ID, derived from snag IDs
snags <- snagMOD %>%
  separate(snagid, c("plotid"), sep = "_", remove=FALSE, extra="drop") %>%
  relocate(plotid, .after = "site")
## Load csv of topo metrics 
tidytopo <- read_csv(file.path(tabs, "snag-trem_REFplots_topo-r25m.csv"))
## Combine all files for use in Random Forest (RF) modeling
snagRF <- full_join(snags, tidytopo, by = "plotid") %>%
  relocate(c(elev, trasp, slope), .after = "z_m") %>%
  drop_na(snagid)
glimpse(snagRF)
##--------------------------------------------------------------------------##
## Inspect summary stats by snag class (by diameter and intactness) ----
##--------------------------------------------------------------------------##
snagRF %>%
    group_by(group) %>%
    summarize(n())
summstat <- snagRF[,c("group","ptd","p1th","p2th","p3th", "n",
                    "GFPmid","LADcv","rumple","ENTmid","VAI","VCI",
                    "elev","trasp","slope")] %>% 
  tidyr::drop_na() %>% 
  group_by(group) %>%
  summarise(mean(n), sd(n)); summstat
##--------------------------------------------------------------------------##
beep(2) #----
```
## *3.2. Set up model parameters*
(RF modeling approach adapted from: 
https://daviddalpiaz.github.io/r4sl/ensemble-methods.html#classification-2)
```{r rf-setup}
##--------------------------------------------------------------------------##
## Setup vars for RF modeling to classify all snags by GROUP ----
##--------------------------------------------------------------------------##
grpVars <- snagRF[,c("group","ptd","p1th","p2th","p3th",
                    "GFPmid","LADcv","rumple","ENTmid","VAI","VCI",
                    "elev","trasp","slope")] %>% 
    rename(c(ptDen = ptd,
             p1stRN = p1th,
             p2ndRN = p2th,
             p3rdRN = p3th
             )) %>%
    tidyr::drop_na(); glimpse(grpVars)
grpVars %>%
    group_by(group) %>%
    summarize(n())
##--------------------------------------------------------------------------##
## Test for model collinearity across predictor variables ----
##--------------------------------------------------------------------------##
rfUtilities::multi.collinear(grpVars[,-1], na.rm=TRUE, perm=TRUE,
                             leave.out=TRUE, n=999, p=0.001)
rfCor <- round(cor(grpVars[,-1], method="pearson", use="all.obs"), 
               digits=2) #Pearson correlations
corrplot(rfCor, method="number", tl.col="black", type="upper")
##--------------------------------------------------------------------------##
## Setup vars for RF modeling split by size class ----
##--------------------------------------------------------------------------##
sizeVars <- snagRF[,c("group","top","ptd","p1th","p2th","p3th",
                    "GFPmid","LADcv","rumple","ENTmid","VAI","VCI",
                    "elev","trasp","slope")] %>% 
    rename(c(ptDen = ptd,
             p1stRN = p1th,
             p2ndRN = p2th,
             p3rdRN = p3th
             )) %>%
    tidyr::drop_na(); glimpse(sizeVars)
##--------------------------------------------------------------------------##
## Split data by size class for RF modeling of small & large snags only ----
##--------------------------------------------------------------------------##
snagSM <- subset(sizeVars, group == "SI" | group == "SB") %>%
  select(-group)
snagLG <- subset(sizeVars, group == "LI" | group == "LB") %>%
  select(-group)
nrow(snagSM); nrow(snagLG) #inspect split by size
##--------------------------------------------------------------------------##
beep(10) #----
##--------------------------------------------------------------------------##
```
## *3.3. Build and test RF model across multiple iterations*
```{r rf-all}
##--------------------------------------------------------------------------##
## Create containers for compiling results of RF model iterations ----
##--------------------------------------------------------------------------##
rfb_grp = list(); imp_grp = list(); prd_grp = list(); cMx_grp = list()
prop_grp_df = data.frame(); imp_grp_df=data.frame(); cM_grp_df = data.frame()
##--------------------------------------------------------------------------##
## Run RF model over 20 iterations ----
##--------------------------------------------------------------------------##
set.seed(588)
for (i in 1:20){
  ## Generate randomly split subsets for model testing and validation
  rf_split_grp <- rsample::initial_split(grpVars, prop = 0.6) #randomly sample
  rf_train_grp <- training(rf_split_grp) #assign training data subset
  rf_vtest_grp <- testing(rf_split_grp)  #assign validation/testing data subset
  prop_train_grp <- prop.table(table(rf_train_grp$group)) #look at groups dist.
  prop_grp_df <- rbind(prop_train_grp, prop_grp_df)
  ## Run RF on training data subset
  rfb_grp[[i]] <- randomForest(group ~ .,
                      data = rf_train_grp, 
                      ntree = 1000,
                      mtry = 13, #set to all preds = bagging (vs rf only)
                      importance = TRUE)
  imp_grp <- importance(rfb_grp[[i]]) #look at variable importance from model
  mir_grp <- round(imp_grp[,6]/max(imp_grp[,6]), digits=2) #MIR for mdGini 
  imp_grp_df <- rbind(imp_grp_df, mir_grp)
  ## Use RF model results to predict classes for testing data subset
  prd_grp[[i]] <- predict(rfb_grp[[i]], newdata = rf_vtest_grp)
  ## Evaluate how well RF model performs classification prediction
  cMx_grp[[i]] <- caret::confusionMatrix(prd_grp[[i]], rf_vtest_grp$group)
  cM_grp_df <- rbind(cM_grp_df, cMx_grp[[i]]$overall)
}
##--------------------------------------------------------------------------##
## Summarize results across iterations into tables ----
##--------------------------------------------------------------------------##
RFnames <- as_tibble(sprintf("iter_%02d", 1:20)) #label iteration number
colnames(prop_grp_df)<-c("prop(SI)", "prop(LI)","prop(SB)", "prop(LB)")
## Comparisons of RF evaluation among iterations
colnames(cM_grp_df)<-c("Accuracy","Kappa","AccLowerCI",
                       "AccUpperCI","AccNull","AccPValue","McNemarPValue")
iterRF_grp <- cbind(RFnames,
                    round(prop_grp_df, digits=2),
                    round(cM_grp_df[,c(1,2,6,7)], digits=3)) %>%
    arrange(desc(Kappa))
head(iterRF_grp)
## Comparisons of standardized variable importance scores among iterations
colnames(imp_grp_df)<-c("ptDen","p1stRN","p2ndRN","p3rdRN","GFPmid","LADcv",
                        "rumple","ENTmid","VAI","VCI","elev","trasp","slope")
mirRF_grp <- cbind(RFnames,imp_grp_df) 
rankMIR_grp <- mirRF_grp[-1] %>%
  mutate_all(as.numeric) %>%
  summarise(across(.cols = everything(), mean))  %>%
  gather() 
rankMIR_grp %>% arrange(desc(value))
##--------------------------------------------------------------------------##
## Inspect "best" RF model iteration (via overall acc & Kappa values) ----
##--------------------------------------------------------------------------##
(grps_best <- rfb_grp[[04]])
(grps_cM <- cMx_grp[[04]])
## Look at variable importance results for best RF model
par(mfrow=c(1,1))
varImpPlot(grps_best, main="Variable Importance Measures for top RF model ")
plot(grps_best)
randomForest::importance(grps_best)[,1:4]
round(importance(grps_best)[,1:4]/max(abs(importance(grps_best)[,1:4])),
      digits=2)  #MIR for classes
##--------------------------------------------------------------------------##
beep(8) #----
##--------------------------------------------------------------------------##
```
## *3.4. Build and test RF model across multiple iterations*
```{r rfb-byDiameter}
##--------------------------------------------------------------------------##
## Subset of small snags only (<40cm DBH) ----
##--------------------------------------------------------------------------##
## Create containers for compiling results of RF model iterations
rfb_sm = list(); imp_sm = list(); prd_sm = list(); cMx_sm = list()
prop_sm_df = data.frame(); imp_sm_df = data.frame(); cM_sm_df = data.frame()
## Run RF model over 20 iterations
set.seed(588)
for (i in 1:20){
  rf_sm_split <- initial_split(snagSM, prop = 0.6)
  rf_sm_train <- training(rf_sm_split)
  rf_sm_vtest <- testing(rf_sm_split)
  prop_sm_train <- prop.table(table(rf_sm_train$top))
  prop_sm_df <- rbind(prop_sm_df, prop_sm_train)
  print(prop.table(table(rf_sm_train$top)))
  rfb_sm[[i]] <- randomForest(top ~ .,
                      data = rf_sm_train, 
                      ntree = 1000,
                      mtry = 14, 
                      importance = TRUE)
  imp_sm <- importance(rfb_sm[[i]])
  mir_sm <- round(imp_sm[,4]/max(imp_sm[,4]),
                        digits=2)
  imp_sm_df <- rbind(imp_sm_df, mir_sm)
  prd_sm[[i]] <- predict(rfb_sm[[i]], newdata = rf_sm_vtest)
  cMx_sm[[i]] <- caret::confusionMatrix(prd_sm[[i]], rf_sm_vtest$top)
  cM_sm_df <- rbind(cM_sm_df, cMx_sm[[i]]$overall)
  print(i)
}
## Summarize results across iterations into tables
RFnames <- as_tibble(sprintf("iter_%02d", 1:20))
## Comparisons of RF evaluation among iterations
colnames(prop_sm_df)<-c("prop(BT)", "prop(INTACT)")
colnames(cM_sm_df)<-c("Accuracy","Kappa","AccLowerCI",
                       "AccUpperCI","AccNull","AccPValue","McNemarPValue")
iterRF_sm <- cbind(RFnames,
                    round(prop_sm_df, digits=2),
                    round(cM_sm_df[,c(1,2,6,7)], digits=3)) %>%
    arrange(desc(Kappa)); head(iterRF_sm)
## Comparisons of standardized variable importance scores among iterations
colnames(imp_sm_df)<-c("ptDen","p1stRN","p2ndRN","p3rdRN","GFPmid","LADcv",
                        "rumple","ENTmid","VAI","VCI","elev","trasp","slope")
mirRF_sm <- cbind(RFnames,imp_sm_df) 
rankMIR_sm <- mirRF_sm[-1] %>%
  mutate_all(as.numeric) %>%
  summarise(across(.cols = everything(), mean))  %>%
  gather() %>% arrange(desc(value)); rankMIR_sm
## Inspect "best" RF model iteration (via overall acc & Kappa values)
(sm_best <- rfb_sm[[15]])
(grps_cM <- cMx_sm[[15]])
importance(sm_best)
importance(sm_best)/max(abs(importance(sm_best)[,1:2]))
varImpPlot(sm_best)
##--------------------------------------------------------------------------##
## Subset of large snags only (≥40cm DBH) ----
##--------------------------------------------------------------------------##
## Create containers for compiling results of RF model iterations
rfb_lg = list(); imp_lg = list(); prd_lg = list(); cMx_lg = list()
prop_lg_df = data.frame(); imp_lg_df = data.frame(); cM_lg_df = data.frame()
## Run RF model over 20 iterations 
set.seed(588)
for (i in 1:20){
  rf_lg_split <- initial_split(snagLG, prop = 0.6)
  rf_lg_train <- training(rf_lg_split)
  rf_lg_vtest <- testing(rf_lg_split)
  prop_lg_train <- prop.table(table(rf_lg_train$top))
  prop_lg_df <- rbind(prop_lg_df, prop_lg_train)
  print(prop.table(table(rf_lg_train$top)))
  rfb_lg[[i]] <- randomForest(top ~ .,
                      data = rf_lg_train, 
                      ntree = 1000,
                      mtry = 14, 
                      importance = TRUE)
  imp_lg <- importance(rfb_lg[[i]])
  mir_lg <- round(imp_lg[,4]/max(imp_lg[,4]),
                        digits=2)
  imp_lg_df <- rbind(imp_lg_df, mir_lg)
  prd_lg[[i]] <- predict(rfb_lg[[i]], newdata = rf_lg_vtest)
  cMx_lg[[i]] <- caret::confusionMatrix(prd_lg[[i]], rf_lg_vtest$top)
  cM_lg_df <- rbind(cM_lg_df, cMx_lg[[i]]$overall)
  print(i)
}
## Summarize results across iterations into tables 
RFnames <- as_tibble(sprintf("iter_%02d", 1:20))
## Comparisons of RF evaluation among iterations
colnames(prop_lg_df)<-c("prop(BT)", "prop(INTACT)")
colnames(cM_lg_df)<-c("Accuracy","Kappa","AccLowerCI",
                       "AccUpperCI","AccNull","AccPValue","McNemarPValue")
iterRF_lg <- cbind(RFnames,
                    round(prop_lg_df, digits=2),
                    round(cM_lg_df[,c(1,2,6,7)], digits=3)) %>%
  arrange(desc(Kappa)); head(iterRF_lg)
## Comparisons of standardized variable importance scores among iterations
colnames(imp_lg_df)<-c("ptDen","p1stRN","p2ndRN","p3rdRN","GFPmid","LADcv",
                        "rumple","ENTmid","VAI","VCI","elev","trasp","slope")
mirRF_lg <- cbind(RFnames,imp_lg_df) 
rankMIR_lg <- mirRF_lg[-1] %>%
  mutate_all(as.numeric) %>%
  summarise(across(.cols = everything(), mean))  %>%
  gather() %>% arrange(desc(value)); rankMIR_lg
## Inspect "best" RF model iteration (via overall acc & Kappa values)
(lg_best <- rfb_lg[[09]])
(grps_cM <- cMx_lg[[09]])
importance(lg_best)
importance(lg_best)/max(abs(importance(lg_best)[,1:2]))
varImpPlot(lg_best)
##--------------------------------------------------------------------------##
## Compare MIR-adj mdGini across all 3 models ----
##--------------------------------------------------------------------------##
MIR_all <- full_join(rankMIR_grp, rankMIR_sm, by = "key") %>%
  full_join(rankMIR_lg, by = "key")
colnames(MIR_all) <- c("pred.var","RF.all","RF.sm","RF.lg")
MIR_all %>% arrange(desc(RF.all))
##--------------------------------------------------------------------------##
beep(5) #----
##--------------------------------------------------------------------------##
```
## *3.5. Visualize the results of the Random Forest*
``` {r rfExplainer-plots}
##--------------------------------------------------------------------------##
## Process graphics for RF model of choice ----
##--------------------------------------------------------------------------##
## Set focal RF model (shown = top model for RF-all)
rfMod <- grps_best
## Make list of file names 
figList = c("1-min-depth","2-mw-imp","3-intrxns","4-pred")
## Set up plots
mdf <- min_depth_distribution(rfMod)
impf <- measure_importance(rfMod)
impvars <- important_variables(impf, k = 5,
                    measures = c("mean_min_depth", "no_of_trees"))
intx <- min_depth_interactions(rfMod, impvars)
## Generate plots of interest
mindepth <- plot_min_depth_distribution(mdf)
mindepth

mwImp <- plot_multi_way_importance(impf, size_measure = "no_of_nodes")
mwImp

intrxns <- plot_min_depth_interactions(intx)
intrxns

predIntrxn <- plot_predict_interaction(rfMod, grpVars, "p3rdRN", "slope")
predIntrxn

plotList = list(mindepth,mwImp,intrxns,predIntrxn)
## Save plots to png; make a separate file for each plot.
# for (i in 1:4) {
#   file_name = file.path(figs, paste("snag-trem_rfbGRP-40cm-report_fig", 
#                                       figList[[i]], ".png", sep=""))
#   png(file_name,
#       width = 800,
#       height = 600)
#   print(plotList[i])
#   dev.off()
# }
##--------------------------------------------------------------------------##
## Compile a report of all graphics ----
##--------------------------------------------------------------------------##
explain_forest(rfMod, interactions = TRUE, data = grpVars,
               path=file.path(figs,
                              "snag-trem_RF-explainer_rfbGRP-40cm.html"))
##--------------------------------------------------------------------------##
## Set up partition-based tree models ----
##--------------------------------------------------------------------------##
(tree.grps = rpart(group ~ ., data = grpVars)); printcp(tree.grps)
plot(tree.grps); text(tree.grps, use.n = TRUE)
####
## Generate classification tree for just the training data
snag_part = rpart(group ~ ., data = rf_train_grp)
rpart.plot(snag_part)
## 
snag_part_pred = predict(snag_part, rf_vtest_grp, type = "class")
table(predicted = snag_part_pred, actual = rf_vtest_grp$group)
calc_acc = function(actual, predicted) {mean(actual == predicted)}
(part_acc = calc_acc(predicted = snag_part_pred, actual = rf_vtest_grp$group))
caret::confusionMatrix(snag_part_pred, rf_vtest_grp$group)
##--------------------------------------------------------------------------##
beep(4) #----
##--------------------------------------------------------------------------##
```
